---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# optimall

`optimall` offers a collection of functions that are designed to streamline the process of optimum sample allocation, specifically under an adaptive, multi-phase approach. Its main functions allow users to:

* Split existing strata into smaller strata based on values or percentiles of other variables.

* Calculate the optimum number of samples to allocate to each stratum in a given study in order to minimize the variance of an estimate of interest. 

* Optimally allocate a fixed number of samples to an ancillary sampling wave based on results from a prior wave. 

When used together, these functions can automate most of the sampling workflow. This vignette will introduce the theoretical framework behind the functions before demonstrating how they can be used to work with sampling data in R.

## Installation

You can install the released version of optimall from [CRAN](https://CRAN.R-project.org) with:

``` r
# install.packages("optimall")
```

And the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("yangjasp/optimall")
```
## Example

Given a dataframe where each row represents one unit, `optimall` can define the stratum each unit belongs to:

```{r example, eval=FALSE}
library(optimall)
data <- split_strata(data = data, strata = "old_strata", 
                     split_var = "variable_to_split_on", 
                     type = "value", split_at = c(1,2))
```

we can then use `optimum_allocation` to calculate the optimum allocation a fixed number of samples to our strata in order to minimize the variance of a variable of interest.

```{r example2, eval=FALSE}
optimum_allocation(data = data, strata = "new_strata", 
                   y = "var_of_interest", nsample = 100)
```

`optimall` offers more functions that streamline adaptive, multi-wave sampling workflows. For a more detailed description, see package vignettes.
